Supermartingales and Termination Analysis of Statistical PCF

### 2021-01-16 LICS21/intro.tex
minor
Reworked introduction. Added conclusion.

\lo{Proving termination of higher-order probabilistic programs with continuous distribution is important but difficult.}

\medskip

Various theorems about probabilistic programs rely on the assumption that the program terminates almost surely. 
One possible approach to prove almost sure termination (AST) is to find some variant on the program state, called \emph{ranking function}, that decreases on average sufficiently quickly that it must at some point reach 0, at which point the program terminates. 
In other words, the program's behavior is used to define an associated (ranking) \emph{supermartingale}. Proof rules based on relating the program state to supermartingales already exist for first-order imperative programs \cite{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15,DBLP:journals/pacmpl/McIverMKK18}. This paper's contribution is to extend this method to a higher-order setting.

@@

Basic (non-random) lambda calculus has the very useful Church-Rosser property, which implies (among other things) that even if execution of a program starts in a different order, it will still reach the same normal form eventually (assuming it does reach a normal form). 
Probabilistic lambda calculus does not have this property, because random choices may be duplicated, and evaluating the same subterm multiple times can yield different results. 
However, with a restricted set of reduction strategies, Church-Rosserness may be regained. 
We introduce a novel addressing scheme for the possible random choices in a program's execution, which ensures that the same random choices are taken at corresponding positions in alternative reduction sequences, so that the same eventual result can be reached. 
This is then used to prove yet another extension to the ranking function theorem, that (\cref{thm:confluent ranking}) ranking functions may be defined with respect to alternative reduction strategies (which in some cases may lead to a considerably simpler execution and ranking function), and (\cref{thm:confluentNoStronger}) rankability in this sense still imples almost sure termination. 
We stress that the confluent trace semantics is of wider interest, and has other possible applications as well, for example in Bayesian inference algorithms.

@@

\subsubsection*{Contributions}

This is the first application of martingales to probabilistic lambda calculus, and the first version of trace semantics that's capable of satisfying the (restricted) Church-Rosser property.

\lo{Ongoing.}

\subsubsection*{Outline}

>>>>

Probabilistic (or randomised) programs have long been recognised as central to the efficient solution of algorithmic problems \cite{Rabin1976}.
More recently, in probabilistic programming \cite{DBLP:conf/icse/GordonHNR14,rainforth2017Automating,vandemeent2018Introduction}, probabilistic programs are used to express generative models whose posterior probability can be computed by general purpose inference engines.
Sampling from continuous distributions is an essential construct of probabilistic programming languages, such as Church \cite{DBLP:conf/uai/GoodmanMRBT08}, Stan \cite{carpenter2017stan}, Anglican \cite{DBLP:conf/pkdd/TolpinMW15}, Gen \cite{cusumano-towner2019Gen}, Pyro \cite{bingham2019Pyro}, Edward \cite{tran2016edward} and Turing \cite{ge2018Turing}.
Another key feature is higher order; in fact some of the most influential probabilistic languages are functional.

In this paper we study a central property of probabilistic programs: \emph{termination}. 
A program is \emph{almost surely terminating} (AST) if a run of it terminates with probability $1$.
When a probabilistic program implements a solution to an algorithmic problem, one naturally requires the computation to terminate with a high probability, usually $1$.
In probabilistic programming, guarantees of AST are equally important. 
Indeed, it is standard for designers and implementors of probabilistic programming systems to regard non-AST programs as defining invalid models, and hence inadmissible \cite{Rainforth2017,DBLP:conf/uai/GoodmanMRBT08}.

Moreover, various theorems about probabilistic programs rely on the assumption that the program terminates almost surely. 
A recent result \cite{MakOPW20} proves that AST programs have density functions that are differentiable almost everywhere.
This is significant for Bayesian machine learning, because the latter property is a precondition for the correctness of some of the most scalable inference algorithms, such as Hamiltonian Monte Carlo \cite{DBLP:conf/aistats/ZhouGKRYW19,Nishimura2020b} and reparameterised gradient variational inference \cite{DBLP:conf/nips/LeeYY18}.

\subsection*{Contributions}

A possible approach to prove almost sure termination (AST) is to find some variant on the program state, called \emph{ranking function}, that decreases on average sufficiently quickly that it must at some point reach 0, at which point the program terminates. 
In other words, the program's behavior is used to define an associated (ranking) \emph{supermartingale}. Proof rules based on relating the program state to supermartingales already exist for first-order imperative programs \cite{DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15,DBLP:journals/pacmpl/McIverMKK18}. This paper's contribution is to extend this method to a higher-order setting.

@@

Basic (non-random) lambda calculus has the very useful Church-Rosser property, which implies (among other things) that even if execution of a program starts in a different order, it will still reach the same normal form eventually (assuming it does reach a normal form). 
Probabilistic lambda calculus does not have this property, because random choices may be duplicated, and evaluating the same subterm multiple times can yield different results. 
However, with a restricted set of reduction strategies, Church-Rosserness may be regained. 
We introduce a novel addressing scheme for the possible random choices in a program's execution, which ensures that the same random choices are taken at corresponding positions in alternative reduction sequences, so that the same eventual result can be reached. 
This is then used to prove yet another extension to the ranking function theorem, that (\cref{thm:confluent ranking}) ranking functions may be defined with respect to alternative reduction strategies (which in some cases may lead to a considerably simpler execution and ranking function), and (\cref{thm:confluentNoStronger}) rankability in this sense still imples almost sure termination. 
The confluent trace semantics is of wider interest, and has other possible applications as well, for example in Bayesian inference algorithms.

@@

\paragraph*{Outline}

### 2021-01-16 LICS21/related.tex
Introduction:ongoing Reformatted last antitone sparse ranking function in appendix
Reworked introduction. Added conclusion.

\lo{TODO. Related work (.75 column). Discuss \cite{DBLP:conf/birthday/BlancLM05}.} \akr{I think that that labelling scheme is not really that related. Would a note like ``This is another scheme for labelling positions in reachable terms, therefore it appears superficially similar, but it's not actually that related.'' actually be worthwhile? I suppose it could at least reassure readers that we're aware of that paper and it isn't the same thing, but I don't know how likely readers are to actually draw the relation in the first place.}

@@

>>>>

\lo{TODO. Related work (.75 column). Discuss \cite{DBLP:conf/birthday/BlancLM05}.} \akr{I think that that labelling scheme is not really that related. Would a note like ``This is another scheme for labelling positions in reachable terms, therefore it appears superficially similar, but it's not actually that related.'' actually be worthwhile? I suppose it could at least reassure readers that we're aware of that paper and it isn't the same thing, but I don't know how likely readers are to actually draw the relation in the first place.}
\lo{On reflection, I agree it is not necessary to discuss this.}

@@

\paragraph*{Conclusions}

We have presented the first application of martingales to probabilistic lambda calculus, and the first version of trace semantics that's capable of satisfying the (restricted) Church-Rosser property.
Though the construction of ranking functions is inherently difficult, using sparse ranking functions, antitone ranking functions and ranking functions w.r.t.~an alternative reduction strategy, we have shown that a great variety of programs can be shown to be AST, including some that are beyond the reach of methods in the literature.
